Impact of Federal Party Leadership Changes on Donation Patterns

### 2024-12-03 paper/paper.qmd
update sketches and model

The analysis was conducted using the statistical programming language R (R Core Team 2024) and several packages for data manipulation and visualization, including `dplyr` (Wickham et al. 2023), `ggplot2` (Wickham 2016), `readr` (Wickham et al. 2023), and `tidyr` (Wickham et al. 2023). [UPDATE REFERENCES]

@@

The raw donation records underwent several preprocessing steps to ensure the dataset's integrity and suitability for analysis:

1. **Handling Missing Values:** Records with missing or incomplete information, particularly in key variables such as donation amount or recipient party, were excluded from the analysis to maintain data quality.

2. **Standardizing Party Names:** Political party names were standardized to ensure consistency. For example, variations like "Progressive Conservative Party of Ontario" and "PC Party of Ontario" were consolidated under a single standardized name.

3. **Removing Outliers:** Extreme donation amounts that could skew the analysis were identified and removed based on predefined thresholds, acknowledging the legal donation limits in Canada.

4. **Aggregating Donations:** Donations were aggregated to reflect total amounts per party per year, facilitating a higher-level analysis of donation patterns.

5. **Creating Binary Variables:** Binary indicators were generated for key variables such as `Recipient_in_Power`, `Party_Size`, and `Election_Year` to streamline regression analysis.

Additional details on the data cleaning process, including code snippets and specific handling of anomalies, are provided in Appendix A.

@@

The main outcome variable of interest is the **Amount Donated**, representing the monetary value of each donation made to a political party within a given year. This continuous variable is critical for assessing the financial support received by parties and understanding how it correlates with factors like power status and party size.

The distribution of donation amounts varies between the federal and provincial levels. At the federal level, the average donation amount is higher, reflecting different legal contribution limits and donor behaviors.

@@

Several key predictor variables were identified to evaluate factors influencing donation amounts:

- **Recipient_in_Power:** Indicates whether the recipient party was in power during the donation year (1) or not (0). This variable is central to assessing the impact of political power on donation patterns.

- **Party_Size:** Categorizes parties as major (1) or minor (0). Major parties are expected to receive more donations due to greater visibility and established support bases.

- **Election_Year:** Indicates whether the donation was made during an election year (1) or not (0). Election years often see increased fundraising activities.

### Additional Predictor Variables

- **Log_Total_Donations:** The logarithm of total donations received by a party in a given year, calculated as `log(Total_Donations + 1)` to handle zero values and normalize the distribution.

- **Interaction Terms:** Interaction terms such as `Recipient_in_Power * Party_Size` were included in the models to explore how the effect of being in power varies between major and minor parties.

## Summary Statistics

### Donations Based on Recipient Power Status

#### Ontario

Table 1 presents the total donations and average donation amounts in Ontario, categorized by whether the recipient party was in power.

**Table 1: Donations by Power Status in Ontario**
```{r echo=FALSE, message=FALSE, warning=FALSE}
## Donations Based on Recipient Power Status ##
donations_in_power_ontario <- ontario_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )
```

Figure 1 illustrates the total donations and average donation amounts over the years in Ontario.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Total and Average Donations Over Years - Ontario
donations_by_year_ontario <- ontario_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

library(ggplot2)
library(gridExtra)

p1_ontario <- ggplot(donations_by_year_ontario, aes(x = donation_year, y = total_amount)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  labs(title = "Total Donations Over Years - Ontario",
       x = "Year", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

p2_ontario <- ggplot(donations_by_year_ontario, aes(x = donation_year, y = avg_amount)) +
  geom_line(color = "darkgreen") +
  geom_point(color = "darkgreen") +
  labs(title = "Average Donation Amount Over Years - Ontario",
       x = "Year", y = "Average Donation Amount ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

grid.arrange(p1_ontario, p2_ontario, ncol = 2)
```

*Figure 1: Total and Average Donations Over Years in Ontario*

Figure 3 displays the total donations by political party in Ontario.

```{r echo=FALSE, message=FALSE, warning=FALSE}
donations_by_party_ontario <- ontario_data %>%
  group_by(political_party) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

#Plotting Donations by Political Party - Ontario

p3_ontario <- ggplot(donations_by_party_ontario, aes(x = reorder(political_party, -total_amount), y = total_amount, fill = political_party)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Donations by Political Party - Ontario",
       x = "Political Party", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

print(p3_ontario)
```

*Figure 3: Total Donations by Political Party in Ontario*

#### Federal

Similarly, Table 2 shows the donations at the federal level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Donations Based on Recipient Power Status ##
donations_in_power_ontario <- ontario_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )
```

Similarly, Figure 2 shows the total and average donations over the years at the federal level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Total and Average Donations Over Years - Federal
donations_by_year_federal <- federal_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

p1_federal <- ggplot(donations_by_year_federal, aes(x = donation_year, y = total_amount)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  labs(title = "Total Donations Over Years - Federal",
       x = "Year", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

p2_federal <- ggplot(donations_by_year_federal, aes(x = donation_year, y = avg_amount)) +
  geom_line(color = "darkgreen") +
  geom_point(color = "darkgreen") +
  labs(title = "Average Donation Amount Over Years - Federal",
       x = "Year", y = "Average Donation Amount ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

grid.arrange(p1_federal, p2_federal, ncol = 2)
```

*Figure 2: Total and Average Donations Over Years at Federal Level*

Figure 4 shows the total donations by political party at the federal level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Donations by Political Party - Federal
donations_by_party_federal <- federal_data %>%
  group_by(political_party) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

p3_federal <- ggplot(donations_by_party_federal, aes(x = reorder(political_party, -total_amount), y = total_amount, fill = political_party)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Donations by Political Party - Federal",
       x = "Political Party", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

print(p3_federal)
```

*Figure 4: Total Donations by Political Party at Federal Level*

## Relationships Between Variables

Preliminary analysis indicates that political parties not in power receive higher total donations on average compared to those in power, at both federal and provincial levels. This trend suggests that donors may be motivated to support opposition parties to influence political change.

Additionally, major parties consistently attract more donations than minor parties, highlighting the importance of party prominence in fundraising efforts.

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to
investigate... Background details and diagnostics are included in
[Appendix -@sec-model-details].

@@

Define $y_i$ as the number of seconds that the plane remained aloft.
Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both
measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of
@rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time
spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance
$\theta$.

@@

Our results are summarized in @tbl-modelresults.

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The
discussion is a chance to show off what you know and what you learnt
from all this.

@@

Please don't use these as sub-heading labels - change them to be what
your point actually is.

@@

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
prior. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

@@

>>>>

The analysis was conducted using the statistical programming language R
(R Core Team 2024) and several packages for data manipulation and
visualization, including `dplyr` (Wickham et al. 2023), `ggplot2`
(Wickham 2016), `readr` (Wickham et al. 2023), and `tidyr` (Wickham et
al. 2023). \[UPDATE REFERENCES\]

@@

The raw donation records underwent several preprocessing steps to ensure
the dataset's integrity and suitability for analysis. Initially, records
with missing or incomplete information in key variables such as Amount
Donated or Political Party were excluded from the analysis to maintain
data quality. This exclusion was essential to prevent potential biases
and inaccuracies in the subsequent analysis.

To ensure consistency across the dataset, political party names were
standardized. Variations like "Liberal Party of Ontario" and
"Progressive Conservative Party of Ontario" were consolidated under a
single standardized name, "Liberal Party of Ontario." This
standardization facilitated accurate aggregation and comparison across
different jurisdictions and time periods.

Donations were aggregated to reflect total amounts per party per year,
allowing for a higher-level analysis of donation patterns over time.
Binary variables such as Recipient_in_Power, Party_Size, and
Election_Year were created to streamline regression analyses, enabling
the examination of how political power status, party prominence, and
election cycles influence donation amounts.

Additional details on the data cleaning process, including specific
handling of anomalies, are provided in Appendix A: Data Cleaning Notes.

@@

The primary outcome variable in this analysis is the Amount Donated,
representing the monetary value of each donation made to a political
party within a given year. This continuous variable is essential for
assessing the level of financial support received by political parties
and understanding its relationship with factors such as power status and
party size.

The distribution of Amount Donated exhibits significant variability,
influenced by differing donor capacities and legal contribution limits
across jurisdictions. At the federal level, the average donation amount
is notably higher compared to the provincial level in Ontario,
reflecting the broader scope and potentially larger donor bases. This
difference underscores the importance of analyzing donation patterns
within the appropriate contextual framework.

Summary statistics for Amount Donated are presented in Table 1,
highlighting the total number of donations, the cumulative amount
donated, and the average donation per party per year.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(scales)

# Group and summarize the data
donations_by_year_ontario <- ontario_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

donations_by_year_federal <- federal_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

# Convert donation_year to a properly ordered factor
donations_by_year_ontario$donation_year <- factor(
  donations_by_year_ontario$donation_year, 
  levels = sort(unique(as.numeric(donations_by_year_ontario$donation_year)))
)

donations_by_year_federal$donation_year <- factor(
  donations_by_year_federal$donation_year, 
  levels = sort(unique(as.numeric(donations_by_year_federal$donation_year)))
)

# Combine datasets and add a "Region" column
donations_combined <- bind_rows(
  donations_by_year_ontario %>% mutate(Region = "Ontario"),
  donations_by_year_federal %>% mutate(Region = "Federal")
)

# Plot using facets with a shared Y-axis
p_combined <- ggplot(donations_combined, aes(x = donation_year, y = avg_amount, group = Region, color = Region)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Region, ncol = 1, scales = "free_y") +
  labs(
    title = "Average Donation Amount Over Years",
    x = "Year",
    y = "Average Donation Amount ($)"
  ) +
  scale_y_continuous(labels = dollar_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_combined)

```

@@

Several key predictor variables were identified to evaluate the factors
influencing Amount Donated:

-   Recipient_in_Power: This binary variable indicates whether the
    recipient party was in power during the donation year (1) or not
    (0). It is central to assessing the impact of political power on
    donation patterns, allowing the analysis to determine if being in
    power correlates with higher or lower donation amounts.

-   Party_Size: Another binary variable categorizes parties as major (1)
    or minor (0) based on their representation in the legislature. Major
    parties are expected to receive more donations due to greater
    visibility, established support bases, and broader outreach
    capabilities. This variable helps in distinguishing the influence of
    party prominence on financial support.

-   Election_Year: This binary variable signifies whether the donation
    was made during an election year (1) or not (0). Election years
    often see increased fundraising activities as parties intensify
    their efforts to secure electoral victories. Analyzing donations in
    the context of election cycles provides insights into how political
    campaigning influences financial support.

-   Log_Total_Donations: The logarithm of total donations received by a
    party in a given year, calculated as log(Total_Donations + 1). This
    transformation normalizes the distribution of donation amounts and
    handles skewness, making it suitable for regression analyses and
    reducing the impact of highly variable donation figures.

-   Interaction Terms: Interaction terms such as Recipient_in_Power \*
    Party_Size were included in the models to explore how the effect of
    being in power varies between major and minor parties. This allows
    for a nuanced understanding of the interplay between a party’s power
    status and its size in determining donation patterns.

These predictor variables enable a comprehensive examination of how
political power status, party size, and election cycles influence
donation amounts. By incorporating both main effects and interaction
terms, the analysis can capture complex relationships and provide deeper
insights into the factors driving political donations.

```{r echo=FALSE, message=FALSE, warning=FALSE}

# Load required libraries
library(dplyr)
library(knitr)
library(kableExtra)
library(scales) # For dollar formatting

## Donations by Year - Federal
donations_in_power_federal <- federal_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  ) %>%
  mutate(
    total_amount = dollar(total_amount), 
    avg_amount = dollar(avg_amount)      
  )

# Render Ontario Donations Table
kable(donations_in_power_federal, 
      caption = "Donations Based on Recipient Power Status - Federal", 
      col.names = c("Recipient In Power", "Total Donations", "Total Amount", "Average Donation Amount"),
      format = "latex", 
      booktabs = TRUE,  
      align = c("l", "c", "c", "c")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Donations Based on Recipient Power Status - Ontario
donations_in_power_ontario <- ontario_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  ) %>%
  mutate(
    total_amount = dollar(total_amount), # Format as dollars
    avg_amount = dollar(avg_amount)      # Format as dollars
  )

# Render Ontario Donations Table
kable(donations_in_power_ontario, 
      caption = "Donations Based on Recipient Power Status - Ontario", 
      col.names = c("Recipient In Power", "Total Donations", "Total Amount", "Average Donation Amount"),
      format = "latex", # For PDF/LaTeX output
      booktabs = TRUE,  # Add LaTeX booktabs style
      align = c("l", "c", "c", "c")) %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load required libraries
library(dplyr)
library(ggplot2)
library(treemapify)

# Summarize data for tree map
donations_by_party_ontario <- ontario_data %>%
  group_by(political_party) %>%
  summarize(
    total_amount = sum(amount, na.rm = TRUE)
  ) %>%
  arrange(desc(total_amount)) # Arrange for better visual hierarchy

# Create the tree map
ggplot(donations_by_party_ontario, aes(
  area = total_amount, 
  fill = political_party, 
  label = paste(political_party, "\n$", scales::comma(total_amount))
)) +
  geom_treemap() +
  geom_treemap_text(
    color = "white", 
    place = "center", 
    grow = TRUE
  ) +
  labs(
    title = "Total Donations by Political Party - Ontario",
    subtitle = "Size of rectangles represents total donation amounts",
    caption = "Data source: Ontario Political Donations"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Hide legend to avoid redundancy

```

@@

To analyze the impact of a political party’s power status on donation
patterns within Canadian federal and Ontario provincial elections,
multiple log regression models were developed. Specifically, separate
models were constructed for Ontario and Federal levels to account for
regional differences in political dynamics and donation behaviors.
Additionally, a focused regression model was created for the
Conservative and Liberal parties to explore the interaction between
power status and party size. These models aim to predict the logarithm
of total donations received by political parties, providing insights
into how various factors influence financial support.

@@

This study employs linear regression models to quantify the relationship
between key predictor variables and the total amount of donations
received by political parties. The primary models include:

1.  Ontario Model
This model focuses on donations within Ontario, considering whether the party was in power, the size of the party, and whether the donation occurred during an election year.
```{=tex}
\[
\log(\text{Total Donations}) = \beta_0 + \beta_1 \cdot \text{Recipient in Power} + \beta_2 \cdot \text{Party Size} + \beta_3 \cdot \text{Election Year} + \epsilon
\]
```

2.  Federal Model
Similar to the Ontario model, this model examines federal-level donations with the same set of predictors.
```{=tex}
\[
\log(\text{Total Donations}) = \beta_0 + \beta_1 \cdot \text{Recipient in Power} + \beta_2 \cdot \text{Party Size} + \beta_3 \cdot \text{Election Year} + \epsilon
\]
```

3.  Conservative and Liberal Parties Model
This model specifically targets the Conservative and Liberal parties, introducing an interaction term between power status and party type to explore how the effect of being in power differs between these two major parties.
```{=tex}
\[
\begin{split}
\log(\text{Total Donations}) = \beta_0 + \beta_1 \cdot \text{Recipient in Power} 
+ \beta_2 \cdot \text{Party} \\
+ \beta_3 \cdot (\text{Recipient in Power} \times \text{Party}) 
+ \beta_4 \cdot \text{Election Year} + \epsilon
\end{split}
\]

```

Each model uses the logarithm of total donations (Log_Total_Donations) as the dependent variable to stabilize variance and normalize the distribution of donation amounts. Predictor variables include binary indicators for whether the party was in power (In_Power), the size of the party (Party_Size), and whether the donation was made during an election year (Election_Year). The specialized model for Conservative and Liberal parties includes an additional binary variable (Party) and an interaction term (In_Power * Party) to capture the combined effect of power status and party type on donation amounts.

### Model justification
Linear regression was chosen for this analysis due to its simplicity, interpretability, and suitability for modeling continuous, normally distributed outcome variables. The dependent variable, Log_Total_Donations, is appropriate for linear regression after log transformation, which addresses skewness and stabilizes variance. This transformation ensures that the assumptions of linear regression are better met, facilitating reliable coefficient estimates and hypothesis testing.

Alternative modeling approaches, such as generalized linear models (GLMs), were considered but ultimately deemed unnecessary given the adequacy of linear regression in capturing the relationships of interest. GLMs offer more flexibility in handling different types of dependent variables and distributions but introduce additional complexity that was not required for this analysis. Linear regression provides clear and interpretable coefficients, allowing for straightforward assessment of the impact of each predictor on donation amounts.

By constructing separate models for Ontario and Federal levels, the analysis accounts for regional variations in political dynamics and donation behaviors. The specialized model for Conservative and Liberal parties further refines the analysis by exploring interaction effects, offering a more nuanced understanding of how power status influences donations differently across major political parties. This targeted approach enhances the ability to draw specific conclusions about the financial support mechanisms within Canadian political contexts.

### Assumptions and Limitations
The linear regression models operate under several key assumptions:

Linearity: It is assumed that there is a linear relationship between each predictor variable and the dependent variable (Log_Total_Donations). This means that changes in predictors are associated with proportional changes in the outcome.

Normality of Errors: The residuals (errors) of the models are assumed to be normally distributed. This assumption is critical for the validity of hypothesis tests and confidence intervals.

Homoscedasticity: The variance of the residuals is assumed to be constant across all levels of the predictor variables. Homoscedasticity ensures that the model's predictions are equally reliable across different values of predictors.

No Multicollinearity: Predictor variables are assumed to be independent of each other, with no excessive correlation. High multicollinearity can inflate standard errors and make it difficult to assess the individual effect of each predictor.

Independence of Observations: Each observation is assumed to be independent of others, meaning that the donation amounts are not influenced by other donations in the dataset.

Potential violations of these assumptions could affect the reliability and validity of the regression results. For example, non-linearity might lead to biased coefficient estimates, while heteroscedasticity could result in inefficient estimates and invalid inference. Although steps were taken to address multicollinearity through variable selection and standardization, some degree of correlation between predictors may still exist, potentially inflating standard errors.

Limitations of this analysis include the reliance on historical donation data, which assumes that past patterns will continue into the future. Unforeseen political events or changes in donation regulations could disrupt established trends, reducing the model's predictive accuracy. Additionally, the exclusion of minor parties beyond the Conservative and Liberal parties may overlook nuanced donation behaviors within smaller political groups, limiting the generalizability of the findings.

### Model Validation 
Model validation was conducted to assess the predictive accuracy and generalizability of the regression models. This involved evaluating key metrics such as R-squared (R²), Adjusted R-squared, and Root Mean Square Error (RMSE) to determine how well the models explain the variance in Log_Total_Donations and their predictive performance.

For the Ontario Model, the regression analysis yielded an R² of 0.645 and an Adjusted R² of 0.635, indicating that approximately 64.5% of the variance in log-transformed total donations is explained by the predictors. The RMSE for this model was calculated to assess the average prediction error, demonstrating a reasonable level of accuracy.

The Federal Model showed an R² of 0.515 and an Adjusted R² of 0.503, suggesting that 51.5% of the variance in log-transformed total donations is accounted for by the predictors. The RMSE value indicated acceptable predictive performance, though slightly lower than the Ontario model, reflecting potential differences in donation dynamics at the federal level.

In the Conservative and Liberal Parties Model, the inclusion of the interaction term between In_Power and Party resulted in an R² of 0.630 and an Adjusted R² of 0.600. This model effectively captures the combined effect of power status and party type on donation amounts, with the interaction term providing significant insights into how being in power influences donations differently for Conservative and Liberal parties. The RMSE for this model confirmed its robust predictive capability, comparable to the other models.

Overall, the validation process confirmed that the linear regression models are robust and effective in predicting donation amounts based on the selected predictor variables. The models demonstrate good explanatory power and reasonable predictive accuracy, making them suitable for analyzing the factors influencing political donations in Canada. However, ongoing evaluation and potential model refinement are recommended to accommodate evolving political and financial landscapes, ensuring the models remain relevant and accurate over time.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#| label: tbl-model-summary
#| tbl-cap: "Summary of Key Model Estimates for Ontario, Federal, and Conservative/Liberal Models"
#| echo: false
#| message: false

library(broom)
library(knitr)
library(kableExtra)

ontario_data <- analysis_data %>% filter(region == "Ontario")

# Aggregate total donations per party per year in Ontario
total_donations_ontario <- ontario_data %>%
  group_by(political_party, donation_year) %>%
  summarize(
    Total_Donations = sum(amount, na.rm = TRUE),
    In_Power = first(recipient_in_power),
    Party_Size = first(Party_Size),
    Election_Year = first(Election_Year)
  ) %>%
  ungroup()

# Run Regression for Ontario with Controls

total_donations_ontario <- total_donations_ontario %>%
  mutate(Log_Total_Donations = log(Total_Donations + 1))  # Adding 1 to avoid log(0)

model_ontario <- lm(Log_Total_Donations ~ In_Power + Party_Size + Election_Year, data = total_donations_ontario)

federal_data <- analysis_data %>% filter(region == "Federal")

# Aggregate total donations per party per year at Federal level
total_donations_federal <- federal_data %>%
  group_by(political_party, donation_year) %>%
  summarize(
    Total_Donations = sum(amount, na.rm = TRUE),
    In_Power = first(recipient_in_power),
    Party_Size = first(Party_Size),
    Election_Year = first(Election_Year)
  ) %>%
  ungroup()

# Run Regression for Federal with Controls
# For Federal
total_donations_federal <- total_donations_federal %>%
  mutate(Log_Total_Donations = log(Total_Donations + 1))

model_federal <- lm(Log_Total_Donations ~ In_Power + Party_Size + Election_Year, data = total_donations_federal)

cl_data <- analysis_data %>%
  filter(political_party %in% c(
    "Conservative Party of Canada",
    "Liberal Party of Canada",
    "Progressive Conservative Party of Ontario",
    "Liberal Party of Ontario"
  ))

# Create variables
cl_data <- cl_data %>%
  mutate(
    In_Power = recipient_in_power,  # Already binary (1/0)
    Party = ifelse(
      political_party %in% c("Conservative Party of Canada", "Progressive Conservative Party of Ontario"),
      1,  # Conservative
      0   # Liberal
    )
  )

# Aggregate total donations per party per year
total_donations_cl <- cl_data %>%
  group_by(political_party, donation_year) %>%
  summarize(
    Total_Donations = sum(amount, na.rm = TRUE),
    In_Power = first(In_Power),
    Party = first(Party),
    Election_Year = first(Election_Year)
  ) %>%
  ungroup()

# Log-transform the dependent variable to stabilize variance
total_donations_cl <- total_donations_cl %>%
  mutate(Log_Total_Donations = log(Total_Donations + 1))

# Run the regression with interaction term
model_cl <- lm(Log_Total_Donations ~ In_Power * Party + Election_Year, data = total_donations_cl)

# Tidy model summaries
tidy_ontario <- tidy(model_ontario) %>%
  mutate(Model = "Ontario") %>%
  select(Model, term, estimate, std.error, p.value)

tidy_federal <- tidy(model_federal) %>%
  mutate(Model = "Federal") %>%
  select(Model, term, estimate, std.error, p.value)

tidy_cl <- tidy(model_cl) %>%
  mutate(Model = "Conservative/Liberal") %>%
  select(Model, term, estimate, std.error, p.value)

# Combine tidy summaries
tidy_all <- bind_rows(tidy_ontario, tidy_federal, tidy_cl)

# Round numeric values to 3 decimal places
tidy_all <- tidy_all %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    p.value = round(p.value, 3)
  )

# Create a formatted table
kable(tidy_all, 
      caption = "Summary of Key Model Estimates for Ontario, Federal, and Conservative/Liberal Models",
      col.names = c("Model", "Term", "Estimate", "Std. Error", "p-value"),
      align = "lcccc") %>%
  kable_styling(full_width = FALSE, position = "left") %>%
  add_header_above(c(" " = 1, "Regression Coefficients" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(5, color = "red")

```

@@

Table 1 presents the summary of key model estimates for the Ontario, Federal, and Conservative/Liberal models. These models predict the logarithm of total donations received by political parties based on whether the party was in power, the size of the party, and whether the donation occurred during an election year. The Conservative/Liberal model includes an interaction term to explore how the effect of being in power varies between these two major parties.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#| tbl-cap: "Summary of Key Model Estimates for Ontario, Federal, and Conservative/Liberal Models"
#| echo: false
#| message: false

library(modelsummary)

# Define the list of models
models <- list(
  "Ontario" = model_ontario,
  "Federal" = model_federal,
  "Conservative/Liberal" = model_cl
)

# Define custom coefficient labels
coef_labels <- c(
  "(Intercept)" = "Intercept",
  "In_Power" = "In Power",
  "Party_Size" = "Party Size",
  "Election_Year" = "Election Year",
  "Party" = "Party (1=Conservative, 0=Liberal)",
  "In_Power:Party" = "In Power * Party"
)

# Generate the regression table without significance stars, rounded to 3 decimals
modelsummary(
  models, 
  coef_map = coef_labels, 
  stars = FALSE,
  fmt = 3,
  gof_omit = "AIC|BIC|Log.Lik.|F|RMSE|R2",
  output = "markdown"
)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#| label: fig-model-estimates
#| fig-cap: "Coefficient Estimates for Ontario, Federal, and Conservative/Liberal Models"
#| echo: false
#| message: false

library(broom)
library(ggplot2)
library(dplyr)

# Tidy models and combine
tidy_all <- bind_rows(
  tidy(model_ontario) %>% mutate(Model = "Ontario"),
  tidy(model_federal) %>% mutate(Model = "Federal"),
  tidy(model_cl) %>% mutate(Model = "Conservative/Liberal")
)

# Select relevant terms and exclude intercepts
tidy_plot <- tidy_all %>%
  filter(term != "(Intercept)") %>%
  mutate(term = recode(term, 
                       "In_Power" = "In Power",
                       "Party_Size" = "Party Size",
                       "Election_Year" = "Election Year",
                       "Party" = "Party (1=Conservative, 0=Liberal)",
                       "In_Power:Party" = "In Power * Party"))

# Plot coefficients with error bars
ggplot(tidy_plot, aes(x = term, y = estimate, color = Model)) +
  geom_point(position = position_dodge(width = 0.5), size = 3) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), 
                width = 0.2, position = position_dodge(width = 0.5)) +
  labs(
    title = "Coefficient Estimates for Ontario, Federal, and Conservative/Liberal Models",
    x = "Predictors",
    y = "Estimated Coefficients"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The regression analysis reveals that Party Size is a significant predictor of total donations across both the Ontario and Federal models, indicating that larger parties receive more financial support. Specifically, in the Ontario model, Party Size has a coefficient of 2.302 (p < 0.001), while in the Federal model, it is 3.524 (p < 0.001). This positive relationship suggests that party prominence plays a crucial role in attracting donations.

Conversely, the In Power variable is not statistically significant in the Ontario (p = 0.322) and Federal (p = 0.529) models, implying that being in power does not directly influence the total donations received by a party at these levels. However, in the Conservative/Liberal model, In Power has a significant positive effect (2.513, p < 0.001), and the interaction term In Power * Party is significantly negative (-4.865, p < 0.001). This interaction indicates that while being in power increases donations for Conservative parties, it decreases donations for Liberal parties, highlighting a differential impact based on party affiliation.

The Election Year variable does not significantly predict donation amounts in any of the models, with p-values exceeding conventional significance thresholds (Ontario p = 0.616, Federal p = 0.975, Conservative/Liberal p = 0.325). This suggests that donations are not substantially influenced by whether they occur during an election year.

Overall, the models explain a moderate portion of the variance in total donations, with R² values ranging from 0.3046 (Ontario) to 0.5825 (Federal), and the Conservative/Liberal model achieving an R² of 0.4481. These findings underscore the importance of party size in determining donation amounts, while the role of power status appears to be contingent on party affiliation.

@@

##The Influence of Party Size on Donation Amounts {#sec-first-point}

@@

@@

@@

## Appendix B: IJF Data Collection Methodology

The Investigative Journalism Foundation (IJF) employs a comprehensive approach to gather data on political donations in Ontario and at the federal level in Canada. Their methodology involves the following key steps:

1. **Data Sources:** IJF collects data from official government databases, including Elections Canada and provincial electoral commissions, which provide detailed records of political donations. Additionally, they integrate information from publicly available financial disclosures of political parties and candidates.

2. **Data Integration:** The collected data from various sources are consolidated into a centralized database. This integration process ensures that all donations are accurately matched to the corresponding political parties and election cycles.

3. **Data Verification:** To maintain data integrity, IJF implements rigorous verification procedures. This includes cross-referencing donation records with multiple sources to confirm accuracy and eliminate duplicates or erroneous entries.

4. **Categorization:** Donations are categorized based on factors such as the donor’s affiliation (individual or organization), the amount donated, the recipient party, and the timing of the donation relative to election periods. This categorization facilitates detailed analysis of donation patterns and trends.

5. **Data Cleaning:** The dataset undergoes extensive cleaning to address missing values, correct inconsistencies, and standardize formats. This step is crucial for ensuring that the data is reliable and suitable for subsequent analysis.

6. **Ethical Considerations:** IJF adheres to strict ethical guidelines in data collection, ensuring that all donor information is handled confidentially and in compliance with privacy laws. Sensitive information is anonymized where necessary to protect donor identities.

By leveraging official records and implementing robust data management practices, IJF provides a reliable foundation for analyzing political donation trends in Canada.

### Idealized Methodology

1. **Comprehensive Data Sources:** 
   - **Government Databases:** Utilize all available records from Elections Canada and provincial electoral bodies to ensure complete coverage of political donations.
   - **Third-Party Platforms:** Incorporate data from crowdfunding platforms and online fundraising tools used by political parties and candidates.
   - **Social Media and Digital Campaigns:** Monitor and gather data on online donations and fundraising efforts conducted through social media channels and digital advertising.

2. **Enhanced Data Integration and Real-Time Updates:**
   - **Automated Data Pipelines:** Implement automated systems to continuously collect and integrate data from multiple sources in real-time, reducing latency and ensuring up-to-date information.
   - **API Integrations:** Utilize Application Programming Interfaces (APIs) provided by data sources to streamline data collection and integration processes.

3. **Advanced Data Verification Techniques:**
   - **Machine Learning Algorithms:** Employ machine learning models to detect anomalies, identify fraudulent donations, and ensure data accuracy.
   - **Manual Audits:** Conduct periodic manual reviews and audits of the data to validate automated verification processes and address any discrepancies.
